---
title: "ica-wrangling"
format: html
---

**For wrangling exercises**

## Warm-Up 

### Example 1 {.unnumbered}
Which verb would help us…

* keep only information about state names, county names, and the 2020 and 2016 Democratic support (not the 2012 results, demographics, etc)
  * select
* get only the data on Minnesota
  * filter
* define a new variable which calculates the change in Democratic support from 2016 to 2020, using dem_pct_20 and dem_pct_16
  * mutate
* sort the counties from highest to lowest Democratic support
  * arrange
* determine the total number of votes cast across all counties
  * summarize


### Example 2: Select Columns {.unnumbered}

```{r}
elections <- read.csv("https://mac-stat.github.io/data/election_2020_county.csv")
library(tidyverse)

```

```{r}
# What's the first argument? The second?
# first argument: dataset
# second argument: relevant columns
select(elections, c(state_name, county_name, dem_pct_20, dem_pct_16))


# with pipe function
elections |> 
  select(state_name, county_name, dem_pct_20, dem_pct_16)
```

### Example 3: Filter Rows {.unnumbered}

```{r}
# without a pipe
filter(elections, state_name == "Minnesota")

# with a pipe
elections |> 
  filter(state_name == "Minnesota")

```

### Example 4: Filter and Select {.unnumbered}

```{r}
# Without pipes
filter(select(elections, c(state_name, county_name, dem_pct_20, dem_pct_16)), state_name == "Minnesota")

# With pipes: all verbs in 1 row (ugly)
elections |> select(state_name, county_name, dem_pct_20, dem_pct_16) |> filter(state_name == "Minnesota")

# With pipes: each verb in a row
elections |>
  select(state_name, county_name, dem_pct_20, dem_pct_16) |>
  filter(state_name == "Minnesota")

# We can even do this with UN-tidyverse code in "base" R
elections[elections$state_name == "Minnesota", c(1, 4, 8, 12)]

# summary: use tidyerse code, the pipe function, and each verb on a new row 
# this is easiest to read
```

### Example 5: Order of Operations

```{r}
# order doesn't matter here
elections |> 
  filter(state_name == "Minnesota") |>
  select(state_name, county_name, dem_pct_20, dem_pct_16)

elections |>
  select(state_name, county_name, dem_pct_20, dem_pct_16) |>
  filter(state_name == "Minnesota")


```
### Example 6: Storing Results {.unnumbered}
```{r}
# Store the results
mn <- elections |> 
  select(state_name, county_name, dem_pct_20, dem_pct_16) |> 
  filter(state_name == "Minnesota")

# Always check it out to confirm it's what you want it to be!
head(mn)
nrow(mn)
nrow(elections)
```
## Exercises

### Exercise 1 {.unnumbered}
```{r}
# create elections_small
elections_small <- elections |>
  select(state_name, county_name, total_votes_20, repub_pct_20, dem_pct_20, total_votes_16, dem_pct_16)

# check it out!
head(elections_small)


elections |>
  filter(dem_pct_16 <23.96)|>
  select(county_name, dem_pct_16)
```


