---
title: "Mulivariate Viz"
---

Use this file for practice with the **mulivariate viz** in-class activity.  Refer to the class website for details.

## Review {.unnumbered}

```{r}
library(tidyverse)

# Import data
weather <- read.csv("https://mac-stat.github.io/data/weather_3_locations.csv") |> 
  mutate(date = as.Date(date))  

# Check out the first 6 rows
head(weather)

# What are the units of observation?
# weather on a certain date

# How many data points do we have? 
dim(weather)
2367*24

# What type of variables do we have?
str(weather)
```

### Example 1 {.unnumbered}

```{r}
ggplot(weather, aes(x=temp3pm))+
  geom_density()
```

### Example 2 {.unnumbered}
How do afternoon temperatures (`temp3pm`) differ by `location`?
```{r}
# plot 2 (no facets or densities)
ggplot(weather, aes(x=temp3pm, fill=location))+
  geom_histogram(color="white")

# plot 3 (facets)
ggplot(weather, aes(x=temp3pm, fill=location))+
  geom_histogram(color="white",binwidth=3)+
  facet_wrap(~location)

```

### Example 3 {.unnumbered}
```{r}
# Don't worry about the syntax (we'll learn it soon)
woll <- weather |>
  filter(location == "Wollongong") |> 
  mutate(date = as.Date(date))  

# How often does it raintoday?
# Fill your geometric layer with the color blue.
ggplot(woll, aes(x = raintoday))+
  geom_bar(fill="blue")

# If it does raintoday, what does this tell us about raintomorrow?
# Use your intuition first
ggplot(woll, aes(x = raintoday, fill=raintomorrow))+
  geom_bar(position="fill") 

```

### Example 4 {.unnumbered}
```{r}
# THINK: What variable goes on the y-axis?
# For the curve, try adding span = 0.5 to tweak the curvature
ggplot(woll, aes(x=date, y=temp3pm))+
  geom_point()+
  geom_smooth(span=0.5)


# Instead of a curve that captures the general TREND,
# draw a line that illustrates the movement of RAW temperatures from day to day
# NOTE: We haven't learned this geom yet! Guess.
ggplot(woll, aes(y = temp3pm, x = date))+
  geom_point()+
  geom_line()
```
* the temperature is higher in January/summer months, but lower in July/winter months
* it tends to vary with the seasons in a sinusoidal pattern
* some outliers (eg very hot days or very cold days)

### Example 5 {.unnumbered}
```{r}
# Plot temp3pm vs temp9am
# Change the code in order to indicate the location to which each data point corresponds
ggplot(weather, aes(y = temp3pm, x = temp9am, color=location)) + 
  geom_point()
  scale_shape_manual(values=c(3,4,2))

# Change the code in order to indicate the location to which each data point corresponds
# AND identify the days on which it rained / didn't raintoday
ggplot(weather, aes(y = temp3pm, x = temp9am, color=location, shape=raintoday)) + 
  geom_point()+
  scale_shape_manual(values=c(1,2))

```
```{r}
# Change the code in order to construct a line plot of temp3pm vs date for each separate location (no points!)
ggplot(weather, aes(y = temp3pm, x = date, color = location)) + 
  geom_line()
```
## In-Class Exercises
```{r}
# Import and check out data
education <- read.csv("https://mac-stat.github.io/data/sat.csv")
head(education)

library(tidyverse)
```

### Exercise 1 {.unnumbered}
```{r}
ggplot(education, aes(x=sat))+
  geom_density()
```
**Observations**
* The main peak is around 900, but there is a second, smaller peak near 1050
* range: 800 to 1100
* the mean might be skewed high due to the high outliers, but expected outcome is between 900 and 1000
* states with less poor people have higher scores...


### Exercise 2 {.unnumbered}
```{r}
# Construct a plot of sat vs expend
# Include a "best fit linear regression model" (HINT: method = "lm")

ggplot(education, aes(x=expend, y=sat))+
  geom_point()+
  geom_smooth(method="lm")

# Construct a plot of sat vs salary
# Include a "best fit linear regression model" (HINT: method = "lm")

ggplot(education, aes(x=salary, y=sat))+
  geom_point()+
  geom_smooth(method="lm")
```
**Observations**
* The relationship between spending (expendature or salary) seems to be negative...
* As spending goes up, SAT scores go down (seems counter-intuitive)
* However, the relationship doesn't look very strong

### Exercise 3 {.unnumbered}
```{r}
ggplot(education, aes(x=salary, y=sat, color=expend))+
  geom_point()
```

### Exercise 4 {.unnumberd}
```{r}
# splits the best-fit line into multiple lines to better fit the data
# changing "2" to "3" generates more lines
ggplot(education, aes(y = sat, x = salary, color = cut(expend, 3))) + 
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm")
```
* for most of the data, there is a negative relationship between salary and SAT
* but, when salary is greater than 45, the relationship is positive. these few points have a higher salary and expenses, but lower SAT scores than states w/ low salary and low spending

### Exercise 5 {.unnumbered}
```{r}
ggplot(education, aes(x=fracCat))+
  geom_bar()
```
```{r}
ggplot(education, aes(x=sat, fill=fracCat))+
  geom_density(alpha=0.7)+
  facet_wrap(~fracCat)
```
* states where a lower fraction of students take the SAT have a higher average SAT score (~1050), but states with a greater number of students that take the SAT have a noticiably lower average score (~900)
* if less students take the test, it's probably the best students who take it, which causes the average to go up
* if more students take the SAT in a state, the greater number of test-takers causes the average to go down (eg people who are forced to take it and don't try)

```{r}
ggplot(education, aes(x=expend, y=sat, color=fracCat))+
  geom_point()+
  geom_smooth(method="lm")
```
* states w/ higher average SAT scores spend the same amount, if not less, than states with lower SAT scores
* however, the cluster of points with high SAT scores are all from states with low percentage of students that take the test (0-15%)
* when just looking at the blue points (45-100%), as spending increases, SAT score does appear to increase a little bit (the same trend is true within each fraction-SAT score goes up as spending increases)
* the high outliers (high SAT, low spending) skew the slope of the best-fit line to be negative 


**Simpson's Paradox**
* Because different states have varying percentages of students that took the SAT, comparing SAT scores and spending shows that as spending increases, SAT score decreases.
* The fraction of students that took the test (fracCat) is a confounding variable and skews the data visualization
* if not required to take the sat, if a student is behind or struggling in school, they probably wouldn't take the SAT and wouldn't bring the average down

## Optional Exercises

