---
title: "Wrangling"
format: html
---

Use this file for practice with the **wrangling** in-class activity. Refer to the class website for details.


## Warm-Up {.unnumbered}

### Example 1 {.unnumbered}
Which verb would help us…

* keep only information about state names, county names, and the 2020 and 2016 Democratic support (not the 2012 results, demographics, etc)
  * select
* get only the data on Minnesota
  * filter
* define a new variable which calculates the change in Democratic support from 2016 to 2020, using dem_pct_20 and dem_pct_16
  * mutate
* sort the counties from highest to lowest Democratic support
  * arrange
* determine the total number of votes cast across all counties
  * summarize


### Example 2: Select Columns {.unnumbered}

```{r}
elections <- read.csv("https://mac-stat.github.io/data/election_2020_county.csv")
library(tidyverse)

```

```{r}
# What's the first argument? The second?
# first argument: dataset
# second argument: relevant columns
select(elections, c(state_name, county_name, dem_pct_20, dem_pct_16))


# with pipe function
elections |> 
  select(state_name, county_name, dem_pct_20, dem_pct_16) |>
  head()
```

### Example 3: Filter Rows {.unnumbered}

```{r}
# without a pipe
filter(elections, state_name == "Minnesota") |>
  head()

# with a pipe
elections |> 
  filter(state_name == "Minnesota") |>
  head()

```

### Example 4: Filter and Select {.unnumbered}

```{r}
# Without pipes
filter(select(elections, c(state_name, county_name, dem_pct_20, dem_pct_16)), state_name == "Minnesota") |>
  head()

# With pipes: all verbs in 1 row (ugly)
elections |> select(state_name, county_name, dem_pct_20, dem_pct_16) |> filter(state_name == "Minnesota") |> head()

# With pipes: each verb in a row
elections |>
  select(state_name, county_name, dem_pct_20, dem_pct_16) |>
  filter(state_name == "Minnesota") |>
  head()

# We can even do this with UN-tidyverse code in "base" R
elections[elections$state_name == "Minnesota", c(1, 4, 8, 12)] |>
  head()

# summary: use tidyerse code, the pipe function, and each verb on a new row 
# this is easiest to read
```

### Example 5: Order of Operations {.unnumbered}

```{r}
# order doesn't matter here
elections |> 
  filter(state_name == "Minnesota") |>
  select(state_name, county_name, dem_pct_20, dem_pct_16) |>
  head()

elections |>
  select(state_name, county_name, dem_pct_20, dem_pct_16) |>
  filter(state_name == "Minnesota") |>
  head()


```
### Example 6: Storing Results {.unnumbered}
```{r}
# Store the results
mn <- elections |> 
  select(state_name, county_name, dem_pct_20, dem_pct_16) |> 
  filter(state_name == "Minnesota") |>
  head()

# Always check it out to confirm it's what you want it to be!
head(mn)
nrow(mn)
nrow(elections)
```
## In-Class Exercises {.unnumbered}

### Exercise 1 {.unnumbered}
```{r}
# create elections_small
elections_small <- elections |>
  select(state_name, county_name, total_votes_20, repub_pct_20, dem_pct_20,
         total_votes_16, dem_pct_16) 

# check it out
head(elections_small)


```
### Exercise 2 {.unnumbered}

```{r}
# Keep only data on counties in Hawaii
elections_small |>
  filter(state_name == "Hawaii") |>
  head()


# What does this do?
# gets data for Hawaii and Delaware counties
elections_small |> 
  filter(state_name %in% c("Hawaii", "Delaware")) |>
  head()

# Keep only data on counties where the Republican got MORE THAN 93.97% of the vote in 2020
# THINK: What variable is relevant here?
elections_small |>
  filter(repub_pct_20 > 93.97) |>
  head()

# Keep only data on counties where the Republican got AT LEAST 93.97% of the vote in 2020
# This should have 1 more row (observation) than your answer above
elections_small |>
  filter(repub_pct_20 >= 93.97) |>
  head()
```

```{r}

# Keep only data on counties in Texas where the Democrat got more than 65% of the vote in 2020
# Do this 2 ways.
# Method 1: 2 filters with 1 condition each
elections_small |>
  filter(state_name == "Texas") |>
  filter(dem_pct_20 > 65) |>
  head()

# Method 2: 1 filter with 2 conditions
elections_small |> 
  filter(state_name == "Texas", dem_pct_20 > 65) |>
  head()

```

### Exercise 3 {.unnumbered}

```{r}
# Arrange the counties in elections_small from lowest to highest percentage of 2020 Republican support
# Print out just the first 6 rows
elections_small |>
  arrange(repub_pct_20)|>
  head()

# Arrange the counties in elections_small from highest to lowest percentage of 2020 Republican support
# Print out just the first 6 rows
elections_small |>
  arrange(desc(repub_pct_20)) |>
  head()

```


### Exercise 4 {.unnumbered}

```{r}
# What did this code do?
# creates a new column with republican pct minus democratic pct
elections_small |> 
  mutate(diff_20 = repub_pct_20 - dem_pct_20) |> 
  head()

# What did this code do?
# gets the number of republican votes (total votes times pct GOP)
elections_small |> 
  mutate(repub_votes_20 = round(total_votes_20 * repub_pct_20/100)) |> 
  head()

# What did this code do?
# creates new column saying if GOP won 
elections_small |> 
  mutate(repub_win_20 = repub_pct_20 > dem_pct_20) |> 
  head()
```
```{r}
# You try

# Define a variable that calculates the change in Dem support in 2020 vs 2016
elections_small |> 
  mutate(diff_20 = dem_pct_20 - dem_pct_16) |> 
  head()

# You try
# Define a variable that determines whether the Dem support was higher in 2020 than in 2016 (TRUE/FALSE)
elections_small |> 
  mutate(dem_diff = dem_pct_20 > dem_pct_16) |> 
  head()
```

### Exercise 5 {.unnumbered}
```{r}
elections_small |> 
  filter(state_name == "Wisconsin",
         repub_pct_20 < dem_pct_20) |> 
  arrange(desc(total_votes_20)) |> 
  head()

# Now try it. Change the order of filter and arrange below.
elections_small |> 
  arrange(desc(total_votes_20)) |> 
  filter(state_name == "Wisconsin",
         repub_pct_20 < dem_pct_20) |> 
  head()


elections_small |> 
  filter(state_name == "Delaware") |> 
  mutate(repub_win_20 = repub_pct_20 > dem_pct_20) |> 
  select(county_name, repub_pct_20, dem_pct_20, repub_win_20) |>
  head()

# switch mutate and select
elections_small |> 
  mutate(repub_win_20 = repub_pct_20 > dem_pct_20) |> 
  filter(state_name == "Delaware") |> 
  select(county_name, repub_pct_20, dem_pct_20, repub_win_20) |>
  head()
```

### Exercise 6 {.unnumbered}
```{r}
# part a
elections_small |>
  filter(state_name == "Minnesota") |>
  select(county_name, dem_pct_20) |>
  arrange(desc(dem_pct_20)) |>
  head()

# part b
mn_wi <- elections_small |>
  filter(state_name %in% c("Minnesota", "Wisconsin")) |>
  mutate(dem_change = dem_pct_20 - dem_pct_16) |>
  select(state_name, county_name, dem_pct_20, dem_pct_16, dem_change) |>
  arrange(dem_change)

head(mn_wi)

# part c
ggplot(mn_wi, aes(x=dem_change, fill=state_name))+
  geom_histogram(binwidth = 1,
                 color = "white")+
  facet_wrap(~state_name)
```

### Exercise 7 {.unnumbered}
```{r}
# What does this do?
# gets the median (middle number) of the repub_pct_20 column
elections_small |> 
  summarize(median(repub_pct_20))

# What does this do?
# renames the column name
elections_small |> 
  summarize(median_repub = median(repub_pct_20))

# What does this do?
# gets the median republican percentage and median number of total votes of all the counties
elections_small |> 
  summarize(median_repub = median(repub_pct_20), 
            total_votes = sum(total_votes_20))
```
### Exercise 8 {.unnumbered}

```{r}
elections_small |> 
  group_by(state_name)

# Check out the structure before and after group_by
elections_small |> 
  class()

elections_small |> 
  group_by(state_name) |> 
  class()

# What does this do?
# (What if we didn't use group_by?)
# gets us the median GOP percentage and total votes for each state
# if we didn't use group_by, it would give us these statistics for all the states
elections_small |> 
  group_by(state_name) |> 
  summarize(median_repub = median(repub_pct_20), 
            total_votes = sum(total_votes_20)) 
```

### Exercise 9 {.unnumbered}
```{r}
# Sort the *states* from the most to least total votes cast in 2020
elections_small |>
  group_by(state_name) |>
  select(state_name, total_votes_20) |>
  summarize(total_votes = sum(total_votes_20)) |>
  arrange(desc(total_votes)) |>
  head()

# In 2020, what were the total number of votes for the Democratic candidate and the total number of votes for the Republican candidate in each *state*?
elections_small |>
  group_by(state_name) |>
  select(state_name, total_votes_20, dem_pct_20, repub_pct_20) |>
  mutate(dem_votes = round(total_votes_20 * dem_pct_20/100),
         repub_votes = round(total_votes_20 * repub_pct_20/100)) |>
  summarize(sum(dem_votes),
            sum(repub_votes)) |>
  head()

# What states did the Democratic candidate win in 2020?
elections_small |>
  group_by(state_name) |>
  select(state_name, total_votes_20, dem_pct_20, repub_pct_20) |>
  mutate(dem_votes = round(total_votes_20 * dem_pct_20/100),
         repub_votes = round(total_votes_20 * repub_pct_20/100)) |>
  summarize(total_dem = sum(dem_votes),
            total_repub = sum(repub_votes)) |>
   mutate(dem_win = total_dem > total_repub)|>
  filter(dem_win == "TRUE") |>
  head()
  
```
### Exercise 10 {.unnumbered}
```{r}
world_cup <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/worldcups.csv")
```

```{r}
# In what years did Brazil win the World Cup?
world_cup |>
  filter(winner == "Brazil") |>
  select(year) |>
  head()

# What were the 6 World Cups with the highest attendance?
world_cup |>
  arrange(desc(attendance)) |>
  select(year, attendance) |>
  head()

```

```{r}
# Construct a univariate plot of goals_scored (no wrangling necessary)
# This provides a visual summary of how the number of goals_scored varies from World Cup to World Cup
ggplot(world_cup, aes(x=goals_scored))+
  geom_histogram(binwidth = 15,
                 color = "white")

# Let's follow up the plot with some more precise numerical summaries
# Calculate the min, median, and max number of goals_scored across all World Cups
# NOTE: Visually compare these numerical summaries to what you observed in the plot
world_cup |>
  summarize(median = median(goals_scored),
            max = max(goals_scored),
            min = min(goals_scored))|>
  head()
```

```{r}
# Construct a bivariate plot of how the number of goals_scored in the World Cup has changed over the years
# No wrangling necessary
ggplot(world_cup, aes(x=year, y=goals_scored))+
  geom_point()+
  labs(title = "Goals over Time")

# Our above summaries might be a bit misleading.
# The number of games played at the World Cup varies.
# Construct a bivariate plot of how the typical number of goals per game has changed over the years

world_cup <- world_cup |>
  mutate(goals_per_game = goals_scored/games)

ggplot(world_cup, aes(x=year, y=goals_per_game))+
  geom_point()+
  labs(title = "Goals per Game over Time")
```
### Exercise 11 {.unnumbered}
```{r}
# import data
water_insecurity<-read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2025/2025-01-28/water_insecurity_2023.csv")
library(socviz)
data(county_map)
library(tidyverse)

# Add 0's at the beginning of any geoid that's fewer than 5 numbers long
water_insecurity <- water_insecurity |> 
  mutate(geoid = as.character(geoid)) |> 
  mutate(geoid = 
           ifelse(nchar(geoid) == 4, paste0("0", geoid), geoid))


# update the data to include counties not in the original dataset
counties_with_water <- county_map |>
  left_join(water_insecurity,by = c("id" = "geoid")) |>
  mutate(percent_lacking_plumbing = (plumbing / total_pop) * 100) |>
  select(long, lat, order, hole, piece, group, id, total_pop, plumbing, percent_lacking_plumbing)
```

