---
title: "Data Import"
format: html
---

# Exercises {.unnumbered}

## Exercise 1 {.unnumbered}

### Part b {.unnumbered}
* the dataset is from Kaggle, but the data itself is from TMBd
* metadata on 5000 movies (ie homepage, id, original title, overview, popularity, production companies, production countries, release date, spoken languages, status, tagline, vote_average)
* generated from The Movie Database API

## Exercise 2 {.unnumbered}
```{r}
# part a: absolute path
library(tidyverse)
imdb_messy <- read_csv("C:\\Users\\sarah\\OneDrive\\Documents\\GitHub\\portfolio-sradz\\data\\imdb_5000_messy.csv")

```
```{r}
# part b
getwd()
```

```{r}
# part c
imdb_temp <- read_csv("..\\data\\imdb_5000_messy.csv")
```

```{r}
# part d
# Try importing only the first 5 rows
read_csv("..\\data\\imdb_5000_messy.csv", n_max = 5)
```
## Exercise 3 {.unnumbered}
```{r}
# part b
imdb_messy |>
  mutate(across(where(is.character), as.factor)) |>  # convert characters to factors in order to summarize
  summary()
```
* What type of info is provided on quantitative variables?
  * Stuff about the movie actors/directors or plot or movie itself: title, color, director, actor names, genre, plot keywords, lagnuage 
What type of info is provided on categorical variables?
  * Info about movie statistics: revenue, rating on IMDB, facebook likes
What stands out to you in these summaries? Is there anything you’d need to clean before using this data?
  * some inconsistencies with spelling or formatting (ie different ways of saying color--COLOR, color, Color)
  * a decent amount of NA's or data missing

## Exercise 4 {.unnumbered}

* the formatting of categories within the variable "color" isn't consistent 
* "color" has 6 categories (ie color, Color, COLOR or B&W, Black and White, and NA)
* the largest is Color, but there are a few "COLOR" or "color" movies

```{r}
imdb_messy |>
  count(color)
```


## Exercise 5 {.unnumbered}

```{r}
# clean using factors
imdb_temp <- imdb_messy |>
  mutate(color = fct_recode(color, 
                            "Color" = "COLOR",
                            "Color" = "color",
                            "Black_White" = "Black and White",
                            "Black_White" = "B&W")) 

imdb_temp |>
  count(color)
```
```{r}
# clean using strings!
imdb_temp <- imdb_messy |>
  mutate(color = str_to_lower(color),
         color = str_replace(color, "black and white", "Black_white"),
         color = str_replace(color, "b&w", "Black_white"),
         color = str_replace(color, "color", "Color"))

imdb_temp |>
  count(color)
```

## Exercise 6 {.unnumbered}
```{r}
# Count the total number of rows in imdb_messy
count(imdb_messy)

# Then count the number of NAs in each column
colSums(is.na(imdb_messy))

# Then count the number of NAs in a specific column
imdb_messy |> 
filter(is.na(actor_1_facebook_likes))
```

## Exercise 7 {.unnumbered}

```{r}
# removed the NA's
imdb_temp |>
  summarize(avg_duration = mean(duration, na.rm=TRUE))
```

```{r}
# drop_na() gets rid of any data entries that contain an "NA"
# gets rid of entries that could be helpful (ie NA for a different variable, but not for the one we care about, would be removed)
imdb_temp <- drop_na(imdb_messy)
nrow(imdb_temp)

```

```{r}
ggplot(imdb_messy, aes(x = duration, fill = color)) + 
  geom_density()
```
```{r}
imdb_temp <- imdb_messy |>
  filter(!is.na(duration), !is.na(color))

nrow(imdb_temp) # now we have 5010 movies! (as opposed to 3756 from above when we used drop_na()


# plot (no grey "NA" graph)
ggplot(imdb_temp, aes(x=duration, fill = color))+
  geom_density()
```
```{r}
imdb_temp <- imdb_messy |>
  mutate(actor_1_facebook_likes = replace_na(actor_1_facebook_likes, 0))

imdb_temp |>
  summarize(sum(is.na(actor_1_facebook_likes)))
```

